Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer

### 2024-02-27 index.qmd
write more scaffolding & about bilateral formalism

---
title: "Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer"
authors:
  - name: Roman Ludwig
    orcid: 0000-0001-9434-328X
    email: roman.ludwig@usz.ch
  - name: Yoel Perez Haas
    email: yoel.perezhaas@usz.ch
  - name: Jan Unkelbach
    orcid: 0000-0002-4275-990X
    email: jan.unkelbach@usz.ch
format:
  html:
    code-fold: true
jupyter: python3
execute: 
  cache: true
---

@@

>>>>

---
title: "Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer"
authors:
  - name: Roman Ludwig
    orcid: 0000-0001-9434-328X
    email: roman.ludwig@usz.ch
  - name: Yoel Perez Haas
    email: yoel.perezhaas@usz.ch
  - name: Jan Unkelbach
    orcid: 0000-0002-4275-990X
    email: jan.unkelbach@usz.ch
format:
  html:
    code-fold: true
jupyter: python3
execute: 
  cache: true
crossref:
  fig-prefix: "figure"
  tbl-prefix: "table"
  eq-prefix: "equation"
  sec-prefix: "section"
---

@@

## Modelling Unilateral Lymphatic Spread

- what we previously did

$$
P \left( \mathbf{X} \mid \mathbf{Z} \right) = \frac{P \left( \mathbf{Z} \mid \mathbf{X} \right) P \left( \mathbf{X} \right)}{P \left( \mathbf{Z} \right)}
$$ {#eq-uni-bayes-law}

Let us for later use define the matrices $\boldsymbol{\Lambda}$:

$$
\boldsymbol{\Lambda} = P \left( \mathbf{X} \mid \mathbf{t} \right) = \begin{pmatrix}
\boldsymbol{\pi}^\intercal \cdot \mathbf{A}^0 \\
\boldsymbol{\pi}^\intercal \cdot \mathbf{A}^1 \\
\vdots \\
\boldsymbol{\pi}^\intercal \cdot \mathbf{A}^{t_\text{max}} \\
\end{pmatrix}
$$ {#eq-lambda-matrix}

The $n$-th row in this matrix corresponds to the distribution over hidden states after $t=n-1$ time-steps.

## Extending the Formalism to Both Sides of the Neck

In analogy to @eq-uni-bayes-law, we can write down the posterior joint probability of an involvement $\mathbf{X}^\text{i}$ ipsilaterally _and_ an involvement $\mathbf{X}^\text{c}$ contralaterally, given a diagnosis of the ipsilateral LNLs $\mathbf{Z}^\text{i}$ and of the contralateral ones $\mathbf{Z}^\text{c}$:

$$
P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \mid \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c} \right) = \frac{P \left( \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c} \mid \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \right) P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \right)}{P \left( \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c} \right)}
$$

The likelihood of the diagnoses given a hidden state simply factorise: $P \left( \mathbf{Z}^\text{i}, \mathbf{Z}^\text{c} \mid \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \right) = P \left( \mathbf{Z}^\text{i} \mid \mathbf{X}^\text{i} \right) \cdot P \left( \mathbf{Z}^\text{c} \mid \mathbf{X}^\text{c} \right)$. And the two factors are contained in the observation matrices $\mathbf{B}^\text{i}$ and $\mathbf{B}^\text{c}$.

The term representing the model's prior probability of hidden involvement does not factorize. However, if we assume that lymphatic spread typically does not cross the mid-sagittal line, we can write it as a factorising sum:

$$
\begin{aligned}
P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \right) &= \sum_{t=0}^{t_\text{max}} p(t) \cdot P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \mid t \right) \\
&= \sum_{t=0}^{t_\text{max}} p(t) \cdot P \left( \mathbf{X}^\text{i} \mid t \right) \cdot P \left( \mathbf{X}^\text{c} \mid t \right)
\end{aligned}
$$

This assumption makes intuitive sense: The two sides of the lymphatic network in a typical patient are approximately mirror images of each other. Thus, no major vessels cross the mid-sagittal line. There may, however, be diffusion of lymph fluid accross this line or bulky involvement that redirects lymphatic drainage significantly.

Using this assumption along with @eq-lambda-matrix, we can write the above distribution algebraically as a product:

$$
P \left( \mathbf{X}^\text{i} = \boldsymbol{\xi}_n, \mathbf{X}^\text{c} = \boldsymbol{\xi}_m \right) = \left[ \boldsymbol{\Lambda}^\intercal_\text{i} \cdot \operatorname{diag} p(\mathbf{t}) \cdot \boldsymbol{\Lambda}^\intercal_\text{c} \right]_{n,m}
$$

## Parameter Symmetries

- spread from LNL to LNL probably symmetric

## Tumors Extending Over the Mid-Sagittal Line

- higher prevalence of contralateral involvement when tumor crosses the mid-sagittal plane
- but still lower than on ipsilateral side
- spread to contralateral side in case of midline extension may be describes as linear combination of ipsi- and contralateral spread in case of no midline extension.
- only one additional "mixing parameter" $\alpha$
- tumor's midline extension modelled as a binary RV (lateralized or crossing/extending)
